Left-linear rules in adhesive categories

### 2024-02-12 icalp/fusione.tex

\begin{definition} 
	Let $\X$ be an  $\mathcal{M}$-adhesive category is of \emph{type $\mathbb{B}$} if:
	\begin{enumerate}
		\item every arrow in $\mathcal{M}$ is a coproduct coprojection;
		\item $\X$ has all pushouts;
		\item $\X$ has strict initial objects and, for every object $X$, the arrow $?_X\colon 0\to X $ belongs to $\mathcal{M}$;
		\item all pushouts are $\mathcal{M}$-stable.
	\end{enumerate}
\end{definition}

@@

\noindent
\parbox{10cm}{\begin{proposition}\label{prop:ext}
Let $\X$ be an $\mathcal{M}$-adhesive category of type $\mathbb{B}$. Then for every diagram as the one aside, in which the bottom row is a coproduct cocone and the vertical arrows are in $\mathcal{M}$, the top row is a coproduct if and only if the two squares are pullbacks.
\end{proposition}}
\parbox{2cm}{\vspace{-.4em}$\xymatrix{A\ar[r]^-{f} \ar@{>->}[d]_{r}& C  \ar@{>->}[d]_{s}& B \ar@{>->}[d]^{t}\ar[l]_-{g}\\X  \ar[r]_-{\iota_X}& X+Y & Y \ar[l]^-{\iota_Y}}$}

@@

The previous result entails the following one, needed to show that  an $\mathcal{M}$-adhesive category of type $\mathbb{B}$ satisfy the $\mathcal{M}$-mixed and $\mathcal{M}$-pushout decomposition properties.

\noindent
\parbox{10.7cm}{\begin{proposition}\label{prop:po2}
		Let $\X$ be an $\mathcal{M}$-adhesive category of type $\mathbb{B}$, and suppose that the square on the right is an $\mathcal{M}$ pullback. Then there exists $E\in \X$, $e\colon E\mto C$ in $\mathcal{M}$, $\phi:E\to B_1$ such that $(C, \{m, e\})$ is a coproduct and $g=f+\phi$. Moreover, such a square is a pushout if and only if $\phi$ is an isomorphism.
\end{proposition}} \parbox{3cm}{\vspace{-.15cm}$\xymatrix{A  \ar[r]^{f} \ar@{>->}[d]_{m}& B_0 \ar@{>->}[d]^{\iota_{B_{0}}}\\ C\ar[r]_-{g} & B_0+B_1}$}
\iffalse 
\begin{proof}
		Pulling back $\iota_{B_0}\colon B_0\to B_0\to B_1$ and $\iota_{B_1}\colon B_1\to B_0 + B_1$ we get the diagram below and the thesis follows from \Cref{prop:ext}. The same diagram allows us to deduce $g=f+\phi$.	
	\parbox{5.2cm}{$\xymatrix{A\ar[r]^-{m} \ar@{>->}[d]_{r}& C  \ar@{>->}[d]_{g}& E \ar@{>->}[d]^{\phi}\ar[l]_-{e}\\B_0  \ar[r]_-{\iota_{B_0}}& B_0+B_1 & B_1 \ar[l]^-{\iota_{B_1}} \\ A & B_0 \\C & B_0+E\\ &&B_0+B_1}$}\parbox{8.8cm}{\vspace{-4.1cm}Let us show the second half of the thesis.

@@

\begin{lemma}\label{lem:prop} Let $\X$ be an $\mathcal{M}$-adhesive category of type $\mathbb{B}$, then $\X$ satisfies the $\mathcal{M}$-mixed and $\mathcal{M}$-pushout decomposition properties.
\end{lemma}
\begin{proof}$\mathcal{M}$-mixed decomposition property. This follows at once from \Cref{prop:stab}.

	\smallskip \noindent  $\mathcal{M}$-pushout-decomposition property.

@@

\begin{proof}
    The first point of \Cref{def:switch} is obviously satisfied. The
    second one follows at once since
    \[m_{\der{E},0}= f_0\circ i_1 \qquad h_{\der{E},1}= g_{1}\circ i_0\]
    Finally, the third point holds because we already know that
    \[m_{\der{E},0}= f_0\circ i_1 \qquad h_{\der{E},1}= g_{1}\circ i_0\]
\end{proof}

>>>>

\begin{definition} 
	Let $\X$ be an  $\mathcal{M}$-adhesive category, the pair $(\X, \mathcal{M})$ is of \emph{type $\mathbb{B}$} if:
	\begin{enumerate}
		\item every arrow in $\mathcal{M}$ is a coproduct coprojection;
		\item $\X$ has all pushouts;
		\item $\X$ has strict initial objects and, for every object $X$, the arrow $?_X\colon 0\to X $ belongs to $\mathcal{M}$;
		\item all pushouts are $\mathcal{M}$-stable.
	\end{enumerate}
\end{definition}

@@

\noindent
\parbox{10cm}{\begin{proposition}\label{prop:ext}
Let $(\X, \mathcal{M})$ be a pair of type $\mathbb{B}$. Then for every diagram as the one aside, in which the bottom row is a coproduct cocone and the vertical arrows are in $\mathcal{M}$, the top row is a coproduct if and only if the two squares are pullbacks.
\end{proposition}}
\parbox{2cm}{\vspace{-.4em}$\xymatrix{A\ar[r]^-{f} \ar@{>->}[d]_{r}& C  \ar@{>->}[d]_{s}& B \ar@{>->}[d]^{t}\ar[l]_-{g}\\X  \ar[r]_-{\iota_X}& X+Y & Y \ar[l]^-{\iota_Y}}$}

@@

The previous result entails the following one, needed to show that  in any pair $(\X, \mathcal{M})$ of type $\mathbb{B}$, the category $\X$ satisfies the $\mathcal{M}$-mixed and $\mathcal{M}$-pushout decomposition properties.

\noindent
\parbox{10.7cm}{\begin{proposition}\label{prop:po2}
		Let $(\X, \mathcal{M})$ be a pair of type $\mathbb{B}$, and suppose that the square on the right is an $\mathcal{M}$ pullback. Then there exists $E\in \X$, $e\colon E\mto C$ in $\mathcal{M}$, $\phi:E\to B_1$ such that $(C, \{m, e\})$ is a coproduct and $g=f+\phi$. Moreover, such a square is a pushout if and only if $\phi$ is an isomorphism.
\end{proposition}} \parbox{3cm}{\vspace{-.2cm}$\xymatrix{A  \ar[r]^{f} \ar@{>->}[d]_{m}& B_0 \ar@{>->}[d]^{\iota_{B_{0}}}\\ C\ar[r]_-{g} & B_0+B_1}$}
\iffalse 
\begin{proof}
		Pulling back $\iota_{B_0}\colon B_0\to B_0\to B_1$ and $\iota_{B_1}\colon B_1\to B_0 + B_1$ we get the diagram below and the thesis follows from \Cref{prop:ext}. The same diagram allows us to deduce $g=f+\phi$.	
	\parbox{5.2cm}{$\xymatrix{A\ar[r]^-{m} \ar@{>->}[d]_{r}& C  \ar@{>->}[d]_{g}& E \ar@{>->}[d]^{\phi}\ar[l]_-{e}\\B_0  \ar[r]_-{\iota_{B_0}}& B_0+B_1 & B_1 \ar[l]^-{\iota_{B_1}} \\ A & B_0 \\C & B_0+E\\ &&B_0+B_1}$}\parbox{8.8cm}{\vspace{-4.1cm}Let us show the second half of the thesis.

@@

\begin{lemma}\label{lem:prop} Let $(\X, \mathcal{M})$ be a pair of type $\mathbb{B}$, then $\X$ satisfies the $\mathcal{M}$-mixed and $\mathcal{M}$-pushout decomposition properties.
\end{lemma}
\begin{proof}$\mathcal{M}$-mixed decomposition property. This follows at once from \Cref{prop:stab}.

	\smallskip \noindent
	\parbox{3cm}{$\xymatrix{X \ar[r]^f \ar@{>->}[d]_{\iota_X}& Y\ar[r]^g \ar@{>->}[d]^{\iota_Y} & Z\ar@{>->}[d]^{\iota_Z}\\X+A  \ar@/_.4cm/[rr]_{(g\circ f) + (\varphi \circ \phi)}\ar[r]^{f+\phi}& Y+B \ar[r]^{g+\varphi}& Z+C}$} \qquad \qquad  \qquad   \qquad \parbox{8cm}{ $\mathcal{M}$-pushout-decomposition property. Using \Cref{prop:ext,prop:po2} and the fact that arrows in $\mathcal{M}$ are coproduct coprojections, we can reduce to prove our for diagrams as the one on the left. By hypothesis and \Cref{prop:po2} $\varphi$ and $\varphi\circ \phi$ are both isomorphisms. therefore $\phi$ is an isomorphism too and we can conclude again using \Cref{prop:po2}.\qedhere 
}

@@

\begin{proof}
    The first point of \Cref{def:switch} is obviously satisfied. The
    second one follows at once since $m_{\der{E},0}= f_0\circ i_1$ and $h_{\der{E},1}= g_{1}\circ i_0$.    Finally, the third point holds because we already know that $m_{\der{E},0}= f_0\circ i_1$ and $h_{\der{E},1}= g_{1}\circ i_0$.
\end{proof}
